---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Set default options and your working directory
```{r setoptions, echo=TRUE}
opts_chunk$set(echo=TRUE)
setwd("~/R Code/RepData_PeerAssessment1")
```

## Load the data in a usable form and display a summary of data
```{r loadprocessdata}
RawData <- read.csv("activity.csv",header = TRUE)
summary(RawData)
```


## What is mean total number of steps taken per day across all intervals?

```{r meansteps}
ProcessData <- na.omit(RawData) ## remove NA's
TotalStepsDay <- aggregate(.~date,data=ProcessData,FUN=sum) ## aggregate by date and sum steps
hist(TotalStepsDay$steps, main = "Total Steps Each Day", xlab = "Steps")
MeanSteps <- mean(TotalStepsDay$steps)
MedianSteps <- median(TotalStepsDay$steps)
```
### The mean number of steps taken is `r sprintf("%0.0f",MeanSteps)`.
### The median number of steps taken is `r sprintf("%0.0f",MedianSteps)`.

## What is the average daily activity pattern?
```{r dailypattern}
AveStepsDay <- aggregate(.~interval,data=RawData,FUN=mean) ## find average steps each day for each by 5 minute interval
with(AveStepsDay,plot(interval,steps,type="l"))
title(main="Average Steps Taken per 5 minute Interval",xlab = "5MinInterval",ylab = "Ave Steps")

MaxIdx <- which.max(AveStepsDay$steps) ## find index of max value
```

### The `r AveStepsDay$interval[MaxIdx]` is the 5 minute interval, on average over all days, that contains the maximum number of steps

## Find and replace the missing values with the average steps taken in that 5 minute interval
```{r missingvalues}
IsNA <- is.na(RawData$steps) ## find the missing values
NumNA <- sum(IsNA)  ## count the total number of missing observations
IdxNA <- which(IsNA)
AveStepsInt <- aggregate(.~interval,data=RawData,FUN=mean) ## ave steps for each interval
NewData <- RawData ## newdata has the NA's filled in
## loop over each NA value
for (i in 1:length(IdxNA)) {
    tempIdx <- which(NewData$interval[IdxNA[i]] == AveStepsInt$interval)
    NewData$steps[IdxNA[i]] <- AveStepsInt$steps[tempIdx]
}
## plot filled in data
TotalStepsDay_New <- aggregate(.~date,data=NewData,FUN=sum) ## aggregate by date and sum steps
hist(TotalStepsDay_New$steps, main = "Total Steps Each Day (No NA's)", xlab = "Steps")
MeanStepsNew <- mean(TotalStepsDay_New$steps)
MedianStepsNew <- median(TotalStepsDay_New$steps)
```
### There are `r NumNA` missing observations.
### The mean number of steps after NA replacement is `r sprintf("%0.0f",MeanStepsNew)`.
### The median number of steps after NA replacement is `r sprintf("%0.0f",MedianStepsNew)`.
### The mean after replacement differs by `r sprintf("%0.0f",MeanSteps - MeanStepsNew)`.
### The median after replacement differs by `r sprintf("%0.0f",MedianSteps - MedianStepsNew)`.

## Are there differences in activity patterns between weekdays and weekends?
